Scheeper's Meager-NWD Game and the Menger Game

### 2014-05-12 phd/dissertation/chapter2.tex
almost finished chapter 2 draft

Within this model, we may imagine two players $\pl A$ and $\pl B$ playing a 
game which consists of \term{rounds} enumerated for each $n<\omega$.
During round $n$, $\pl A$ chooses $a_n\in M$, followed by $\pl B$ choosing
$b_n\in M$. The playthrough corresponding to those choices would be
the sequence $p = \<a_0,b_0,a_1,b_1,\dots\>$. If $p\in W$, then $\pl A$
is the winner of that playthrough, and if $p\not\in W$, then $\pl B$ is
the winner. (No ties are allowed.)

@@

As an illustration, we could model a game of chess (ignoring stalemates) 
by letting
\[
  M = \{\<p,s\>: p \text{ is a chess piece and } 
                 s \text{ is a space on the board } \}
\]
representing moving a piece $p$ to the space $s$ on the board. Then the
rules of chess restict White from moving pieces which belong to
Black, or moving a piece to an illegal space on the board. 
\footnote{
  In practice, $M$ is often defined as the union of two sets, such
  as white pieces and black pieces in chess. For example,
  the first player may choose open sets in a topology, while the second player
  chooses points within the topological space.
}
The winning condition could then
``inspect'' the resulting positions of pieces on the board after each move 
to see if White attained a
checkmate. This winning condition along with the rules implicitly define the 
set $W$ of winning playthroughs for White.

@@

\begin{game}
  Let $\bakergame{A}$ denote a game with players $\pl A$ and $\pl B$,
  defined for each subset $A\subset \mathbb{R}$.
  In round $0$, $\pl A$ chooses a number $a_0$, followed by $\pl B$ choosing
  a number $b_0$ such that $a_0<b_0$.
  In round $n+1$, $\pl A$ chooses a number $a_{n+1}$ such that 
  $a_n<a_{n+1}<b_n$, followed by $\pl B$ choosing a number $b_{n+1}$ such that
  $a_{n+1}<b_{n+1}<b_n$.

  $\pl A$ wins the game if the sequence $\<a_n:n<\omega\>$ converges to a 
  point in $A$, and $\pl B$ wins otherwise.
\end{game}

@@

Topological games are usually infinite games for non-trivial spaces.
(The meaning of trivial here depends on the game played.)
One of the earliest examples
of a topological game is the Banach-Mazur game, proposed by Stanislaw Mazur
as Problem 43 in Stefan Banach's Scottish Book (1935). A more comprehensive
history of the Banacy-Mazur and other topological games may be found in 
Telgarsky's survey on the subject \cite{telgarsky}.

@@

  If a winning strategy exists for player $\pl A$ in the game $G$, then we
  write $\pl A \win G$. Otherwise, we write $\pl A \not\win G$.
\end{defn}

Of course, a strategy $\sigma$ is not a winning strategy for a player if there 
exists some \term{counter-attack} by the opponent for which the result is 
a victory for the opponent. Typically this counter-attack is defined in
terms of the strategy $\sigma$; else, the counter-attack is itself a
winning strategy depending on only the round number (which we will
investigate further in a later section).

\begin{defn}
  A game $G$ with players $\pl A$, $\pl B$ is said to be \term{determined}
  if either $\pl A \win G$ or $\pl B \win G$.
  Otherwise, the game is \term{indetermined}.
\end{defn}

\begin{thm}\cite{gale}
  If the move set $M$ for a game $G=\<M,W\>$ is given the discrete topology
  and $W$ is either an open or closed subset of $M^\omega$ with the
  usual product topology, then $G$ is determined.
\end{thm}

This is actually a special case of the powerful Borel Determinacy theorem 
which states that $G=\<M,W\>$ is determined whenever $W$ is a Borel subset
of $M^\omega$. \cite{martin}

It's an easy corollary that all finite games are determined
($W$ must be clopen). As stated earlier, most topological games are infinite,
and many are indetermined for certain spaces constructed using the Axiom of
Choice.
\footnote{
  The Axiom of Choice is required, as the Axiom of Determinacy stating
  that all Gale-Stewart games are determined is another axiom independent
  of ZF. \cite{MR0140430}
}

\begin{ex}
  Let $B$ be a Bernstein subset of the real line. Then $\bmgame{B}$ is
  undetermined.
\end{ex}

\begin{proof}
  It can be shown that $B$ is a Baire space, and we will soon see that 
  $\pl E\not\win \bmgame{X}$ characterizes Baire spaces. Also, if $\sigma$
  is a winning strategy for $\pl N$ in $\bmgame{Y}$ for $Y\subseteq X$, it can 
  be shown that $Y$ contains a closed uncountable set in $X$. 
  Thus $\pl N\not\win \bmgame{B}$.
\end{proof}

@@

The presence or absence of a winning strategie for a player in a topological
game characterizes a property of the topological space in question.

A classical result follows.

@@

\section{Examples of Topological Games}

(TODO: )

>>>>

Within this model, we may imagine two players $\pl A$ and $\pl B$ playing a
game which consists of \term{rounds} enumerated for each $n<\omega$.
During round $n$, $\pl A$ chooses $a_n\in M$, followed by $\pl B$ choosing
$b_n\in M$. The playthrough corresponding to those choices would be
the sequence $p = \<a_0,b_0,a_1,b_1,\dots\>$. If $p\in W$, then $\pl A$
is the winner of that playthrough, and if $p\not\in W$, then $\pl B$ is
the winner. Note that no ties are allowed.

@@

As an illustration, we could model a game of chess (ignoring stalemates)
by letting
\[
  M = \{\<p,s\>: p \text{ is a chess piece and }
                 s \text{ is a space on the board } \}
\]
representing moving a piece $p$ to the space $s$ on the board. Then the
rules of chess restict White from moving pieces which belong to
Black, or moving a piece to an illegal space on the board.\footnote{
  In practice, $M$ is often defined as the union of two sets, such
  as white pieces and black pieces in chess. For example,
  the first player may choose open sets in a topology, while the second player
  chooses points within the topological space.
}
The winning condition could then
``inspect'' the resulting positions of pieces on the board after each move
to see if White attained a
checkmate. This winning condition along with the rules implicitly define the
set $W$ of winning playthroughs for White.

@@

\begin{game}
  Let $\bakergame{X}$ denote a game with players $\pl A$ and $\pl B$,
  defined for each subset $A\subset \mathbb{R}$.
  In round $0$, $\pl A$ chooses a number $a_0$, followed by $\pl B$ choosing
  a number $b_0$ such that $a_0<b_0$.
  In round $n+1$, $\pl A$ chooses a number $a_{n+1}$ such that
  $a_n<a_{n+1}<b_n$, followed by $\pl B$ choosing a number $b_{n+1}$ such that
  $a_{n+1}<b_{n+1}<b_n$.

  $\pl A$ wins the game if the sequence $\<a_n:n<\omega\>$ limits to a
  point in $X$, and $\pl B$ wins otherwise.
\end{game}

@@

Topological games are usually infinite games, ignoring trivial examples.
One of the earliest examples
of a topological game is the Banach-Mazur game, proposed by Stanislaw Mazur
as Problem 43 in Stefan Banach's Scottish Book (1935). A more comprehensive
history of the Banacy-Mazur and other topological games may be found in
Telgarsky's survey on the subject \cite{telgarsky}.

@@

  If a winning strategy exists for a player $\pl A$ in the game $G$, then we
  write $\pl A \win G$. Otherwise, we write $\pl A \not\win G$.
\end{defn}

To show that a winning strategy exists for a player (i.e. $\pl A\win G$),
we typically begin by
defining it and showing that it is \term{legal}: it only yields moves which are
legal according to the rules of the game. Then, we consider an arbitrary
legal attack, and prove that the result of the game is a victory for that
player.

If we wish to show that a winning strategy does not exist for a player
(i.e. $\pl A\not\win G$), we
often consider an arbitrary legal strategy, and use it to define a legal
\term{counter-attack} for the opponent. If we can prove that the result of
the game for that strategy and counter-attack is a victory for the opponent,
then a winning strategy does not exist.

Unlike finite games, is not the case that a winning strategy must exist for
one of the players in an infinite game.

\begin{defn}
  A game $G$ with players $\pl A$, $\pl B$ is said to be \term{determined}
  if either $\pl A \win G$ or $\pl B \win G$.
  Otherwise, the game is \term{undetermined}.
\end{defn}

The Borel Determinacy Theorem states that $G=\<M,W\>$ is determined whenever
$W$ is a Borel subset of $M^\omega$ \cite{martin}. It's an easy corollary that
all finite games are determined; $W$ must be clopen.

However, as stated earlier, most topological games are infinite,
and many are undetermined for certain spaces constructed using the Axiom of
Choice.\footnote{
  These spaces cannot be constructed just only the axioms of ZF. In fact,
  mathematicians have studied an Axiom of Determinacy which declares that
  that all Gale-Stewart games are determined (and implies that the Axiom
  of Choice is false). \cite{MR0140430}
}

@@

The power of studying these infinite-length games can be illustrated by
considering the following proposition.

\begin{prop}
  If $X$ is countable, then $\pl B \win \bakergame{X}$.
\end{prop}

\begin{proof}
  Adapted from \cite{baker}.
  Let $X=\{x_0,x_1,\dots\}$. Let $i(a,b)$ be the least integer such that
  $a<x_{i(x,y)}<b$, if it exists. We define a strategy $\sigma$ for $\pl B$
  such that:
  \begin{itemize}
    \item $\sigma(\<a_0\>)=x_{i(a_0,\infty)}$.
          If $i(a_0,\infty)$ does not exist, then the choice of $\sigma(\<a_0\>)$
          is arbitary, say, $a_0+1$.
    \item $\sigma(\<a_0,...,a_{n+1}\>)=x_{i(a_{n+1},b_n)}$, where
          $b_n=\sigma(\<a_0,\dots,a_n\>)$.
          If $i(a_{n+1},b_n)$ does not exist, then the choice of
          $\sigma(\<a_0,...,a_{n+1}\>)$ is arbitrary, say,
          $
            \sigma(\<a_0,...,a_{n+1}\>)
              =
            \frac{a_{n+1}+b_n}{2}
          $.
  \end{itemize}

  Observe that $\sigma$ is a legal strategy according to the rules of the game
  since $a_0<\sigma(\<a_0\>)$ and $a_{n+1}<\sigma(\<a_0,\dots,a_{n+1}\>)<b_n$.
  We claim this is a winning strategy for $\pl B$. Let $a=\<a_0,a_1,\dots\>$
  be a legal attack by $\pl A$ against $\sigma$: we will show that the
  resulting playthrough is a victory for $\pl B$, that is,
  $\lim_{n\to\infty} a_n\not\in X$. Let $b_n=\sigma(\<a_0,\dots,a_n\>)$.
  Note that
    \[
      a_0<a_1<\dots<\lim_{n\to\infty}a_n<\dots<b_1<b_0
    \]

  If $i(a_0,\infty)$ does not exist, then $a_0$ is greater than every element
  of $X$, and thus
  $\lim_{n\to\infty} a_n \not\in X$. A similar argument follows if some
  $i(a_{n+1},b_n)$ does not exist.

  Otherwise,
    \[
      i(a_0,\infty)<i(a_1,b_0)<i(a_2,b_1)<\dots
    \]
  and for each $i<\omega$, one of the following must hold.
    \begin{itemize}
      \item $i<i(a_0,\infty)$. Then $x_i\leq a_0<\lim_{n\to\infty}a_n$.
      \item $i = i(a_0,\infty)$. Then $x_i=b_0>\lim_{n\to\infty}a_n$.
      \item $i(a_0,\infty)<i<i(a_1,b_0)$.
            Then $x_i\leq a_1<\lim_{n\to\infty}a_n$ or
            $x_i\geq b_0>\lim_{n\to\infty}a_n$.
      \item $i = i(a_{n+1},b_n)$ for some $n<\omega$.
            Then $x_i=b_{n+1}>\lim_{n\to\infty}a_n$.
      \item $i(a_{n+1},b_n)<i<i(a_{n+2},b_{n+1})$ for some $n<\omega$.
            Then $x_i\leq a_{n+2}<\lim_{n\to\infty}a_n$ or
            $x_i\geq b_{n+1}>\lim_{n\to\infty}a_n$.
    \end{itemize}
  In any case, $x_i\not=\lim_{n\to\infty}a_n$, and thus
  $\lim_{n\to\infty}a_n\not\in X$.
\end{proof}

This yields a classical result.

\begin{cor}
  $\mathbb{R}$ is uncountable.
\end{cor}

\begin{proof}
  $\pl A\win \bakergame{\mathbb{R}}$, since $a_n$ must converge to some
  real number. This implies $\pl B\not\win \bakergame{\mathbb{R}}$, and thus
  $\mathbb{R}$ is not countable.
\end{proof}

Infinite games thus provide a rich framework for considering questions in
set theory and topology. In general, the presence or absence of a winning
strategy for a player in a topological game characterizes a property of the
topological space in question.

@@

So far we have assumed both players enjoy \term{perfect information}, and
may develop strategies which use all of the previous moves of the opponent
as input.

\begin{defn}
  For a game $G=\<M,W\>$, the \term{$k$-tactical fog-of-war} is the function
  $\nu_k:M^{<\omega}\to M^{\leq k}$ defined by
    \[
      \nu_k(\<m_0,\dots,m_{n-1}\>)=\<m_{n-k},\dots,m_{n-1}\>
    \]
  and the \term{$k$-Mark\"ov fog-of-war} is the function
  $\mu_k:M^{<\omega}\to (M^{\leq k}\times\omega)$ defined by
    \[
      \mu_k(\<m_0,\dots,m_{n-1}\>)=\<\<m_{n-k},\dots,m_{n-1}\>,n\>
    \]
\end{defn}

Essentially, these fogs-of-war represent a limited memory: $\nu_k$ filters out
all but the last $k$ moves of the opponent, and $\mu_k$ filters out all but
the last $k$ moves of the opponent and the round number.

We call strategies which do not require full recollection of the opponent's
moves \term{limited information strategies}.

\begin{defn}
  A \term{$k$-tactical strategy} or \term{$k$-tactic} is a limited information
  strategy of the form $\sigma\circ\nu_k$.

  A \term{$k$-Mark\"ov strategy} or \term{$k$-mark} is a limited information
  strategy of the form $\sigma\circ\mu_k$.
\end{defn}

$k$-tactics and $k$-marks may then only use the last $k$ moves of the opponent,
and in the latter case, also the round number.

The $k$ is usually omitted when $k=1$. A ($1$-)tactic is
called a \term{stationary strategy} by some authors. $0$-tactics are not
usually interesting (such strategies would be constant functions); however,
we will discuss $0$-Mark\"ov strategies, called
\term{predetermined strategies} since such a strategy only uses the round
number and does not rely on knowing which moves the opponent will make.

\begin{defn}
  If a winning $k$-tactical strategy exists for a player $\pl A$ in the game
  $G$, then we write $\pl A \ktactwin{k} G$. If $k=1$, then $\pl A \tactwin G$.

  If a winning $k$-Mark\"ov strategy exists for a player $\pl A$ in the game
  $G$, then we write $\pl A \kmarkwin{k} G$. If $k=1$, then $\pl A \markwin G$,
  and if $k=0$, then $\pl A \prewin G$.
\end{defn}

The existence of a winning limited information strategy can characterize a
stronger property than the property characterized by a perfect information
strategy.

\begin{defn}
  $X$ is an \term{$\alpha$-favorable space}
  when $\pl N \tactwin \bmgame{X}$.
  $X$ is a \term{weakly $\alpha$-favorable space}
  when $\pl N \win \bmgame{X}$.
\end{defn}

\begin{obs}
  $X$ is $\alpha$-favorable
    $\Rightarrow$
  $X$ is weakly $\alpha$-favorable
    $\Rightarrow$
  $X$ is Baire
\end{obs}

Those arrows may not be reversed. A Bernstein subset of the real line is an
example of a Baire space which is not weakly $\alpha$-favorable, and
